Four decades of water quality change in the upper San Francisco Estuary

### 2017-01-10 sftrends_manu.Rnw
better reporting of censored data

Multi-decadal time series of nutrients and flow records were used to develop a quantitative description of nitrogen trends in the Delta.  The \ac{iep} is a consortium of state and federal agencies that have maintained the \ac{emp} in the Delta region since 1975 \cite{IEP13}.  The \ac{emp} collects monthly water quality samples at 19 stations in the Delta, Suisun Bay, and northeastern San Pablo Bay.  Water samples were collected using a Van Dorn sample, a submersible pump, or a flow through system depending on site.  All samples were processed with standard QA/QC at the California Department of Water Resources Bryte Laboratory in Sacramento \cite{IEP13}.  Nutrient time series were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) at ten discrete sampling stations from 1976 through 2013 (\cref{fig:delt_map}).  Stations were grouped by location in the study area for comparison: Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; middle stations D19, D26, D28; and Suisun stations D4, D6, and D7.  These stations were chosen based on continuity of the water quality time series and geographic location for understanding trends.  Time series were complete for all stations except for an approximate ten year gap from 1996-2014 for D19.  Data were minimally processed with the exception of averaging replicates that occurred on the same day.  Detection limits throughout the period of record were obtained from the metadata, although few observations were censored ($<$ 3\%).  The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two).

@@

obs_sm <- group_by(delt_dat, Site_Code)
@

The observed time series for the ten Delta stations had substantial variation in scale among the nitrogen analytes and differences in apparent seasonal trends (\cref{fig:obsdat}).  In general, long-term (inter-annual) trends were not easily observed from the raw data.

@@

toplo <- delt_dat %>% 
  select(Site_Code, Date, nh, no23) %>% 
  rbind(., missvals) %>% 
  gather('var', 'val', nh:no23) %>% 
  mutate(Site_Code = factor(Site_Code, 
    levels = c('D7', 'D6', 'D4', 'D28', 'D26', 'D19', 'MD10', 'P8', 'C10', 'C3'))
    )

>>>>

% censored percentages
<<echo = F, cache = T>>=
data(mods)
cvals <- mutate(mods, 
  cper = map(data, function(x){

    x <- !na.omit(x$not_cens)
    round(100 * sum(x)/length(x), 1)

  })) %>% 
  select(Site_Code, resvar, cper) %>% 
  unnest %>% 
  arrange(-cper) %>% 
  filter(cper > 3)
@

Multi-decadal time series of nutrients and flow records were used to develop a quantitative description of nitrogen trends in the Delta.  The \ac{iep} is a consortium of state and federal agencies that have maintained the \ac{emp} in the Delta region since 1975 \cite{IEP13}.  The \ac{emp} collects monthly water quality samples at 19 stations in the Delta, Suisun Bay, and northeastern San Pablo Bay.  Water samples were collected using a Van Dorn sample, a submersible pump, or a flow through system depending on site.  All samples were processed with standard QA/QC at the California Department of Water Resources Bryte Laboratory in Sacramento \cite{IEP13}.  Nutrient time series were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) at ten discrete sampling stations from 1976 through 2013 (\cref{fig:delt_map}).  Stations were grouped by location in the study area for comparison: Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; middle stations D19, D26, D28; and Suisun stations D4, D6, and D7.  These stations were chosen based on continuity of the water quality time series and geographic location for understanding trends.  Time series were complete for all stations except for an approximate ten year gap from 1996-2014 for D19.  Data were minimally processed with the exception of averaging replicates that occurred on the same day.  The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two). Less than 3\% of all observations were left-censored, although variation was observed between analytes and location.  The most censored observations were observed for ammonium time series at sites \Sexpr{cvals$Site_Code[1]} (\Sexpr{cvals$cper[1]}\%), \Sexpr{cvals$Site_Code[2]} (\Sexpr{cvals$cper[2]}\%), \Sexpr{cvals$Site_Code[3]} (\Sexpr{cvals$cper[3]}\%), \Sexpr{cvals$Site_Code[4]} (\Sexpr{cvals$cper[4]}\%), and \Sexpr{cvals$Site_Code[5]} (\Sexpr{cvals$cper[5]}\%).

@@

data(mods)

obs_sm <- select(delt_dat, Site_Code, Location, Date, nh, no23, din) %>% 
  gather('var', 'val', nh:din) %>% 
  group_by(Location, var) %>% 
  summarise(val = mean(val, na.rm = T))

The observed time series for the ten Delta stations had substantial variation in scale among the nitrogen analytes and differences in apparent seasonal trends (\cref{fig:obsdat}).  In general, long-term (inter-annual) trends were not easily observed from the raw data.  Total \ac{din} for most stations was dominated by nitrite/nitrate whereas ammonium was present at lower concentrations.  However, C3 had a majority of \ac{din} composed of ammonium and others had larger concentrations ammonium relative to nitrite/nitrate depending on the time of year (e.g., P8, D26).  By location, observed concentrations decreased on average from the upper Delta (C3, C10, MD10, P8), to the middle (D19, D26, D28), and to the Suisun stations(D4, D6, D7).

@@

toplo <- delt_dat %>% 
  select(Site_Code, Date, nh, no23) %>% 
  rbind(., missvals) %>% 
  gather('var', 'val', nh:no23) %>% 
  mutate(Site_Code = factor(Site_Code, 
    levels = rev(c('D7', 'D6', 'D4', 'D28', 'D26', 'D19', 'P8', 'MD10', 'C10', 'C3')))
    )

### 2017-01-10 sftrends_manu.tex
better reporting of censored data

Multi-decadal time series of nutrients and flow records were used to develop a quantitative description of nitrogen trends in the Delta.  The \ac{iep} is a consortium of state and federal agencies that have maintained the \ac{emp} in the Delta region since 1975 \cite{IEP13}.  The \ac{emp} collects monthly water quality samples at 19 stations in the Delta, Suisun Bay, and northeastern San Pablo Bay.  Water samples were collected using a Van Dorn sample, a submersible pump, or a flow through system depending on site.  All samples were processed with standard QA/QC at the California Department of Water Resources Bryte Laboratory in Sacramento \cite{IEP13}.  Nutrient time series were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) at ten discrete sampling stations from 1976 through 2013 (\cref{fig:delt_map}).  Stations were grouped by location in the study area for comparison: Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; middle stations D19, D26, D28; and Suisun stations D4, D6, and D7.  These stations were chosen based on continuity of the water quality time series and geographic location for understanding trends.  Time series were complete for all stations except for an approximate ten year gap from 1996-2014 for D19.  Data were minimally processed with the exception of averaging replicates that occurred on the same day.  Detection limits throughout the period of record were obtained from the metadata, although few observations were censored ($<$ 3\%).  The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two).

@@

The observed time series for the ten Delta stations had substantial variation in scale among the nitrogen analytes and differences in apparent seasonal trends (\cref{fig:obsdat}).  In general, long-term (inter-annual) trends were not easily observed from the raw data.

>>>>

% censored percentages

Multi-decadal time series of nutrients and flow records were used to develop a quantitative description of nitrogen trends in the Delta.  The \ac{iep} is a consortium of state and federal agencies that have maintained the \ac{emp} in the Delta region since 1975 \cite{IEP13}.  The \ac{emp} collects monthly water quality samples at 19 stations in the Delta, Suisun Bay, and northeastern San Pablo Bay.  Water samples were collected using a Van Dorn sample, a submersible pump, or a flow through system depending on site.  All samples were processed with standard QA/QC at the California Department of Water Resources Bryte Laboratory in Sacramento \cite{IEP13}.  Nutrient time series were obtained from the \ac{iep} website (\url{http://water.ca.gov/bdma/meta/Discrete/data.cfm}) at ten discrete sampling stations from 1976 through 2013 (\cref{fig:delt_map}).  Stations were grouped by location in the study area for comparison: Delta stations C3 (Sacramento inflow), C10 (San Joaquin inflow), MD10, P8; middle stations D19, D26, D28; and Suisun stations D4, D6, and D7.  These stations were chosen based on continuity of the water quality time series and geographic location for understanding trends.  Time series were complete for all stations except for an approximate ten year gap from 1996-2014 for D19.  Data were minimally processed with the exception of averaging replicates that occurred on the same day.  The three nitrogen analytes that were evaluated were ammonium, nitrite/nitrate, and \ac{din} (as the sum of the former two). Less than 3\% of all observations were left-censored, although variation was observed between analytes and location.  The most censored observations were observed for ammonium time series at sites C10 (25.4\%), D28 (17.8\%), D19 (12\%), D7 (7.9\%), and D4 (6.4\%).

@@

The observed time series for the ten Delta stations had substantial variation in scale among the nitrogen analytes and differences in apparent seasonal trends (\cref{fig:obsdat}).  In general, long-term (inter-annual) trends were not easily observed from the raw data.  Total \ac{din} for most stations was dominated by nitrite/nitrate whereas ammonium was present at lower concentrations.  However, C3 had a majority of \ac{din} composed of ammonium and others had larger concentrations ammonium relative to nitrite/nitrate depending on the time of year (e.g., P8, D26).  By location, observed concentrations decreased on average from the upper Delta (C3, C10, MD10, P8), to the middle (D19, D26, D28), and to the Suisun stations(D4, D6, D7).
